Fast Sorting on Modern Computers

### 2005-08-02 docs/tech_report/conclusions.tex
fixed small errors. didnt really get rid of any TODOs

No results for the branch prediction performance of sorting algorithm yet
exists. We were able to find some interesting results.

@@

The best stable sort is radixsort. The best comparison-based stable sort is
double-aligned tiled mergesort. The best in-place sort is quicksort. The best in-place
stable sort is, sadly, insertion sort. The best elementary sort is also
insertion sort; as such it should be only used for very small data sets.  The
best sort for very large data sets is radixsort. The best overall sort was
quicksort, though radixsort is very close, and the radixsort considered here was
not optimized. It is expected that an optimized radixsort would beat quicksort
for smaller sets.
% TODO redo above paragraph

>>>>

No results for the branch prediction performance of sorting algorithm yet
exists. We discovered some interesting new results.

@@

The best stable sort is radixsort. The best comparison-based stable sort is
double-aligned tiled mergesort. The best in-place sort is quicksort. The best in-place
stable sort is, sadly, insertion sort. The best elementary sort is also
insertion sort; as such it should be only used for very small data sets.  The
best sort for very large data sets is radixsort. The fastest overall sort was
memory-tuned radixsort, though quicksort is very close.
% TODO redo above paragraph

### 2005-08-02 docs/tech_report/elementary.tex
fixed small errors. didnt really get rid of any TODOs

\displaymath{H_N = 1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{N} =  \sum_{k=1}^{N} \frac{1}{k},  \hspace{0.4cm}  N \geq 1}

@@

\code{Shakersort code}{shakersort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
The instruction count of shakersort is expected to be exactly the same as
bubblesort. The version of shakersort used here, however, does not incorporate
the improvement to end the sort early if the array is already sorted. This will
probably slightly reduce the instruction count as the sort is unlikely to end
early for random keys.

The cache performance of shakersort is expected to be superior to that of
bubblesort. A small amount of temporal reuse occurs at the ends of the array,
as each key is used twice once its loaded into the cache. In addition, when the
array is not more than twice the size of the cache, then the keys in the centre
of the array are not ejected from the cache at all.

>>>>

$$H_N = 1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{N} =  \sum_{k=1}^{N}
\frac{1}{k},  \hspace{0.4cm}  N \geq 1$$

@@

\code{Shakersort code}{shakersort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
The instruction count of shakersort is expected to be exactly the same as
bubblesort, while its cache performance is expected to be superior. A small
amount of temporal reuse occurs at the ends of the array, as each key is used
twice once its loaded into the cache. In addition, when the array is not more
than twice the size of the cache, then the keys in the centre of the array are
not ejected from the cache at all.

### 2005-08-02 docs/tech_report/heap.tex
fixed small errors. didnt really get rid of any TODOs

To take advantage of the heap structure, it must be possible to easily traverse
the heap, going between parent and child, without the presence of links. This
can be done due to the nature of the indices of parents and their children. In
the case of an array indexed between $0$ and $N-1$, a parent whose index is
$i$ has two children with indices of $2i+1$ and $2i+2$. Conversely, the index of
a parent whose child has an index of $j$ is $\lfloor{}(j+1)/2-1\rfloor{}$. It is
also possible to index the array between $1$ and $N$, in which case the children
have indices of $2i$ and $2i+1$, while the parent has an index of
$\lfloor{}j/2\rfloor{}$. Using this scheme can reduces instruction cost, but
results is slightly more difficult pointer arithmetic, which cannot be performed
in some languages\footnote{In C this can be done by \cc{Item a[] =
\&array[-1];}}.

@@

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches logn}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions due to the combination of the shallower heap and the larger
fan-out.

@@

The branch prediction behaves differently to the expected results, as the
prediction ratio gets worse as the fanout gets larger. The reason we predicted
differently was due to selection sort's $O(log N)$ mispredictions; however, this
is over the entire length of the array, and it is not accurate to extrapolate
these results over a sequence only seven keys long.

@@

\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{}{base and memory-tuned heapsorts}

In order to explore the behaviour of the individual branches in
heapsort, we performed a second set of branch prediction
experiments. Rather than using SimpleScalar, we instead added
additional statements to the heapsort source code to simulate a
separate bimodal predictor for each branch. This allowed us to measure
each branch individually, whereas SimpleScalar gives only a total
figure for all branches. An additional advantage is that the resulting
compiled simulation is much faster than a full microarchitecture
simulator, and we were thus able to take the average of thousands of
runs (\textbf{TODO find out exact number}), rather than the small
number of runs possible with SimpleScalar. For these experiments,
randomly generated array of size \textbf{xxxxx} were used.

\textbf{[David] I am confused by this chart, because the labelling is
a bit unclear. There appears to be a base heapsort, a 4-way cache
heapsort and an 8-way cache heapsort. But the labelling doesn't tell
us this.}

@@

\textbf{[David] We need to find some way to describe what each of these
branches is. I don't know myself, so the general reader really, really
won't know.}

>>>>

To take advantage of the heap structure, it must be possible to easily traverse
the heap, going between parent and child, without the presence of links. This
can be done due to the nature of the indices of parents and their children. In
the case of an array indexed between $0$ and $N-1$, a parent whose index is
$i$ has two children with indices of $2i+1$ and $2i+2$. Conversely, the index of
a parent whose child has an index of $j$ is $\lfloor{}(j+1)/2-1\rfloor{}$. It is
also possible to index the array between $1$ and $N$, in which case the children
have indices of $2i$ and $2i+1$, while the parent has an index of
$\lfloor{}j/2\rfloor{}$. Using this scheme can reduces instruction cost, but
results in slightly more difficult pointer arithmetic, which cannot be performed
in some languages\footnote{In C this can be done by \cc{Item a[] =
\&array[-1];}}.

@@

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
3 %TODO does this mean that knuths harmonics applies to this sort of prediction
% thing TODO
mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches logn}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions due to the combination of the shallower heap and the larger
fan-out.

@@

The branch prediction behaves differently to the expected results, as the
prediction ratio gets worse as the fanout gets larger. The reason we predicted
differently was due to selection sort's $O(H_N)$ mispredictions; however, this
is over the entire length of the array, and it is not accurate to extrapolate
these results over a sequence only seven keys long.

@@

\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{}{base
and memory-tuned heapsorts}
%TESTING TODO
TODO move to testing
In order to explore the behaviour of the individual branches in
heapsort, we performed a second set of branch prediction
experiments. Rather than using SimpleScalar, we instead added
additional statements to the heapsort source code to simulate a
separate bimodal predictor for each branch. This allowed us to measure
each branch individually, whereas SimpleScalar gives only a total
figure for all branches. An additional advantage is that the resulting
compiled simulation is much faster than a full microarchitecture
simulator, and we were thus able to take the average of thousands of
runs (\textbf{TODO find out exact number}), rather than the small
number of runs possible with SimpleScalar. For these experiments,
randomly generated array of size \textbf{xxxxx} were used.

@@

%TODO
\textbf{[David] We need to find some way to describe what each of these
branches is. I don't know myself, so the general reader really, really
won't know.}

### 2005-08-02 docs/tech_report/method.tex
minor changes
fixed small errors. didnt really get rid of any TODOs

% TODO maybe remoe this?
Several steps were necessary to prepare for the research.  A framework was
required by which sorts could be tested as they were being developed. This is
discussed in Section \ref{Main}. It was also necessary to fully understand the
technologies to be used in the project. Several texts were used for this: cache
and branch predictors are both in \cite{HennessyPatterson90}; branch predictors
in \cite{Uht97}, as well as in \cite{McFarling93} and \cite{Smith81}; sorts are
discussed in \cite{Sedgewick02} and \cite{Knuth98}; cache-conscious sorts are
discussed \cite{LaMarca96}, \cite{LaMarca99} and \cite{LaMarcaHeap96}.

@@

\subsection{PapiEx}
\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
uses the PAPI framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/~mucci/papiex/}.

@@

LaMarca's final step was to create a multi-mergesort. This attempted to reduce
the number of cache misses in the final merge steps of tiled mergesort by
merging all the sorted arrays in one step. This required the use of an 8-heap,
which had been developed as part of heapsort.
%TODO this isnt what happened exactly

Once these sorts were written it was attempted to improve them slightly.
Instead of a tiled mergesort, a double-aligned tiled mergesort was written.
This attempted to completely sort in the level 1 cache, then to completely sort
in the level 2 cache. Double-aligned multi-mergesort was a combination of this
and of multi-mergesort. The rewrite was based on the base mergesort, and was
written far more cleanly, being more readable and efficient. These were also
found to reduce the instruction count, as well as the level 2 cache miss count.
These results are discussed in more detail in Section \ref{Mergesort results}.

>>>>

Several steps were necessary to prepare for the research.  A framework was
required by which sorts could be tested as they were being developed. This is
discussed in Section \ref{Main}. It was also necessary to fully understand the
technologies to be used in the project. Several texts were used for this: cache
and branch predictors are both in \cite{HennessyPatterson90}; branch predictors
in \cite{Uht97}, as well as in \cite{McFarling93} and \cite{Smith81}; sorts are
discussed in \cite{Sedgewick02} and \cite{Knuth98}; cache-conscious sorts are
discussed \cite{LaMarca96}, \cite{LaMarca99} and \cite{LaMarcaHeap96}.

@@

\subsection{PapiEx}
\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
uses the PAPI framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/\textasciitilde{}mucci/papiex/}.

@@

LaMarca's final step was to create a multi-mergesort. This attempted to reduce
the number of cache misses in the final merge steps of tiled mergesort by
merging all the sorted arrays in one step. This required the use of an k-ary
heap, which had been developed as part of heapsort.

%TODO this bit also has to be checked once we fix mergesort. if w fix mergesort
Once these sorts were written it was attempted to improve them slightly.
Instead of a tiled mergesort, a double-aligned tiled mergesort was written.
This attempted to completely sort in the level 1 cache, then to completely sort
in the level 2 cache. Double-aligned multi-mergesort was a combination of this
and of multi-mergesort. The rewrite was based on the base mergesort, and was
written far more cleanly, being more readable and efficient. These were also
found to reduce the instruction count, as well as the level 2 cache miss count.
These results are discussed in more detail in Section \ref{Mergesort results}.

### 2005-08-02 docs/tech_report/quick.tex
added testing comments
fixed small errors. didnt really get rid of any TODOs

In the case that the array to be sorted fits in the cache, then quicksort has
excellent cache properties. Once it no longer fits in the cache, however, the
number of misses increases dramatically. To fix this problem, LaMarca uses a
technique similar to that used in multi-mergesort. Instead of multi-merging a
series of sorted lists at the end, a \n{multi-way partition} is done at the start,
moving all keys into cache sized containers which can be sorted individually.

@@

\subsection{Test Parameters}

The threshold chosen was 10, simply as this is the number used in Sedgewick's
book.

@@

Despite the sequentially search multi-quicksort performing better than its
binary searched cousin, and its good cache performance, it still failed to
approach the performance of either base or tuned quicksort. In fact, base
quicksort was the fastest of all the quicksorts in our performance counter test,
even though our simulator results show that it has more level 1 misses, level 2
misses, and branch misses than tuned quicksort, and the same number of
instructions.

@@

\sixplot{base_quicksort__median_of_3__0.eps}{base_quicksort__pseudo_median_of_9__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{base,
cache and multi quicksorts and base quicksort (pseudo-median-of-9)}

@@

The memory-tuned quicksort is the only algorithm here that requires a sentinel
check. The branch predictor will only mispredict this once, so the instruction
count increase is only that of removing a single check which occurs relatively
few ($N / THRESHOLD$)\footnote{This is a minimum. The figure may be higher since
partitions are likely to be smaller than the threshold.} times. Still, the
problem is interesting and has applications in other algorithm designs.

>>>>

In the case that the array to be sorted fits in the cache, then quicksort has
excellent cache properties. This applies equally whether discussing a level 1
cache or a level 2 cache. Once the array is larger than the cache, however, the
number of misses increases dramatically. To fix this problem, LaMarca uses a
technique similar to that used in multi-mergesort. Instead of multi-merging a
series of sorted lists at the end, a \n{multi-way partition} is done at the
start, moving all keys into cache sized containers which can be sorted
individually.

@@

\subsection{Test Parameters}
%TESTING
The threshold chosen was 10, simply as this is the number used in Sedgewick's
book.

@@

Despite the sequential search multi-quicksort performing better than its
binary searched cousin, and its good cache performance, it still failed to
approach the performance of either base or tuned quicksort. In fact, base
quicksort was the fastest of all the quicksorts in our performance counter test,
even though our simulator results show that it has more level 1 misses, level 2
misses, and branch misses than tuned quicksort, and the same number of
instructions.

@@

\sixplot{base_quicksort__median_of_3__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{}{base,
cache and multi quicksorts and base quicksort (pseudo-median-of-9)}

@@

The memory-tuned quicksort is the only algorithm here that requires a sentinel
check. The branch predictor will only mispredict this once, so the instruction
count increase is only that of removing a single check which occurs relatively
few ($N / THRESHOLD$) times\footnote{This is a minimum. The figure may be higher since
partitions are likely to be smaller than the threshold.}. Still, the
problem is interesting and has applications in other algorithm designs.

### 2005-08-02 docs/tech_report/radix.tex
fixed small errors. didnt really get rid of any TODOs

\section{Cache Radixsort}
Many improvements exist to radixsort, and some are suggested by Sedgewick and
LaMarca. Some of these improvements help to lower cache misses and instruction
count, both of which slow down radixsort significantly. LaMarca suggests an
improvement by Friend\footnote{See \ref{TODO friend[8]}.} that while the current
byte is being put in place, the counting of the next byte can be done. This
increases the memory requirements, but this is not significant.

@@

Due to the reduced copying, memory-tuned radixsort should have a lower instruction
count and fewer cache misses that base radixsort. There should be nine passes
over the array, leading to just over one miss per key, as opposed to two misses
per key with base radixsort. The complexity and branch prediction results
should remain the same. The instruction count should, however, reduce by nearly
a half to $\frac{9}{16}^ths$ of base radixsort.

>>>>

\section{Memory-tuned Radixsort}
Many improvements exist to radixsort, and some are suggested by Sedgewick and
LaMarca. Some of these improvements help to lower cache misses and instruction
count, both of which slow down radixsort significantly. LaMarca suggests an
improvement by Friend\footnote{See \ref{TODO friend[8]}.} that while the current
byte is being put in place, the counting of the next byte can be done. This
increases the memory requirements, but this is not significant.

@@

Due to the reduced copying, memory-tuned radixsort should have a lower
instruction count and fewer cache misses that base radixsort. There should be
nine passes over the array, leading to just over one miss per key, as opposed to
two misses per key with base radixsort. The complexity and branch prediction
results should remain the same. The instruction count should, however, reduce by
nearly a half to $\frac{9}{16}^{th}s$ of base radixsort.
